How to improve your relationship with your future self, V 2.0

### 2016-10-11 workflow.Rmd
Anonymous version of the MS. We should be able to ignore this branch except in case reviewers have line/page specific edits.

---
title: How to improve your relationship with your future self, V 2.0
author:
- name: Jake Bowers
  affiliation: <jwbowers@illinois.edu>,  University of Illinois at Urbana-Champaign, Departments of Political Science and Statistics; Fellow, White House Social and Behavioral Sciences Team.
- name: Maarten Voors
  affiliation: <maarten.voors@wur.nl>, Wageningen University, Development Economics Group.  \emph{Acknowledgements:} Many thanks to the EGAP Learning Days 2016 participants in Santiago Chile, to the BITSS team and the Department of Economics and Ted Miguel at UC Berkeley where Maarten was a visiting researcher during spring 2016. Maarten gratefully acknowledges financial support from N.W.O. grant 451-14-001. The previous version of this paper benefited from comments and discussions with Mark Fredrickson, Brian Gaines, Kieran Healy, Kevin Quinn, Cara Wong, Mika LaVaque-Manty and Ben Hansen. The source code for this document may be freely downloaded and modified from <https://github.com/jwbowers/workflow>.
bibliography: references.bib
csl: american-political-science-review.csl
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    pdf_document:
      keep_tex: true
      number_sections: true
      fig_width: 7
      fig_height: 7
      fig_caption: true
      template: bowersarticletemplate.latex
      pandoc_args: [
      "-M", "secPrefix=\\S",
      "--filter", "/usr/local/bin/pandoc-crossref"
      ]
      md_extensions: +autolink_bare_uris+ascii_identifiers+tex_math_single_backslash
...

@@

If we cannot count on our memories about why we made such and such a design or analysis decision, then what should we do? How can we minimize regret with our past decisions? How can we improve our relationship with our future self?  This essay is a heavily revised and updated version of @bowers2011tpm and provides some suggestions for practices that will make reproducible data analysis easy and quick. Specifically, this piece aims to amplify some of what we already ought to know and do, and highlight some current practices, platforms, and possibilities.^[@king1995replication and @nagler1995coding were two of the first pieces introducing these kinds of ideas to political scientists. Now, the efforts to encourage transparency and ease of learning from the data and analyses of others has become institutionalized with the DA-RT inititiave <http://www.dartstatement.org/> @lupia2014openness. These ideas are spreading beyond political science as well @freese2007replication and @asendorpf2013recommendations, <http://osf.io> and <http://www.bitss.org/>.] We aim to provide practical advice about how to *do work* such that one complies with such recommendations as a matter of course and, in so doing, can focus personal regret on bad past decisions that do not have to do with data analysis and the production of scholarly papers.

@@

Literate programming need not go against the principle of modular data analysis
[@nagler1995coding]. Jake routinely uses several different files that fulfill
different functions, some of them create \LaTeX code that he can `\input` into
his `main.tex` file, others setup the data, run simulations, or allow him to
record his journeys down blind alleys. Of course, when we have flying cars
running on autopilot, perhaps something other than combining R and Markdown or
LaTeX will make our lives even easier. Then we'll change.

@@

Best practice is to create a file system where you separate folders by function and separate input from output files. For example, below we show a folder structure for a paper where we look at inequality. The paper is written in .tex, data analysis in Stata (that Maarten prefers, so we see .do and .dta file extensions) in R (the program of choice for Jake, see the .R extensions). Maarten likes to make sure that the directory structure in his projects is the same across projects (so, for example, the numbers on the directory names allow for easy default sorting). The file naming with full dates also allows for easy sorting:

@@

Jake tends to be less organized than Maarten and relies on README.md files to tell his future self what is going on in a given directory, a [version control system](http://github.com) to keep track of versions of files, and a [Makefile](http://kbroman.org/minimal_make/) to keep track of relationships between files.

@@

Many rely on Dropbox, 'track changes' in Word, or Google Docs for
collaboration. Those tools tend to require you to communicate with other folks
in your group before you edit existing files. On Dropbox, only one of you can
edit and save a given file at a time. Tracking changes with Word tends to
function best if two people are not editing the same document (merging the
changes between two documents is not simple or builtin). Google Docs allows for
two people to edit the same document at the same time: but you must be online
with a fairly fast internet connection.^[Note to future selves: change this
section every five years or so. Stop saying 'internet' in 2026.] Communication
and trading off editing a file prevents your work (or your colleagues' work)
from getting lost when you both try to save the same file on top of each other. One may also use Dropbox as a kind of server for version
control (though the downside is if your collaborations or project grows large
(in terms of MBs) you need to switch to Dropbox's paid service). Maarten has been using
this for most of his collaborations. You can copy files from the Dropbox
directory into a local working directory so as to avoid clobbering and then
work on merging changes by hand before copying back to the Dropbox directory
and replacing existing files. You will however need to agree on an iron file
and directory naming rule with your collaborators to ensure that the files and directories are always named the same across machines and versions.

Another approach to organizing work on many files, including files for code and
for writing, is to use a formal [version
control](https://en.wikipedia.org/wiki/Version_control) strategy. As of 2016,
the standard for managing large collaborations with many files is the
[Git](https://en.wikipedia.org/wiki/Git_(software)) system. User friendly free
front-ends for Git currently make the process of learning and using Git very
convenient and integrated into the kind of workflow that your future self will
be proud of: [Open Science Framework](http://osf.io)
[bitbucket](http://bitbucket.com), [github](http://github.com). Jake uses
[github](http://github.com) for all of his collaborations with others as well
as with his future self: his collaborators appreciate some of the nice extra
features of github, such as the ability to keep a [shared task
list](https://guides.github.com/features/issues/). Learning about version
control systems takes a bit of time. We suggest, however, it is well worth the
time investment as it will save (lots!) of time later on.

@@

(1) We wrote this paper in the [R markdown](http://rmarkdown.rstudio.com/)
literate programming format using two different text editors (Jake used vim)
and (Maarten used [atom](https://atom.io/)). We organized our files and
collaboration on Github. The source code for this document can be accessed,
downloaded and modified from <https://github.com/jwbowers/workflow>. We also
used Github Issues to send notes to each other and maintain a task list. The
nice thing about github is that is enables you to make "releases" of a project
which enable you to smooth the reproduction of all of the products of your
research (simulations, data analyses, tables, figures, etc..).  Similar
features are offered by the [Open Science Framework (OSF)](https://osf.io/).
Maarten recently participated in a workshop on research transparency organized
by the [Berkeley Institute for Transparency in the Social
Sciences](www.bitss.org) in which all [course material](https://osf.io/qh2nr/)
was accessible through OSF.  OSF integrates with Github nicely and a key
benefit is that both systems have been created to remain for a long time
(unlike perhaps your university drive).  (2) For one paper Jake simply mixed R
and \LaTeX code in one document (called the "Sweave" format of literate
programming) and then added that document and data into into a compressed
archive [@bowers2005dataverse]; (3) For another more computing intensive paper
Jake assembled a set of files that enabled reproduction of results using the
`make` system [@bowers2008dataverse]; (4) Jake also tried the "compendium"
approach [@gentleman2005reproducible,@gentleman2007statistical] which embeds an
academic paper within the R package system [@bowers2011dataverse]; The benefit
of the compendium approach is that one is not required to have access to a
command line for `make`: the compendium is downloadable from within R using
`install.packages()` and is viewable using the `vignette()` function in any
operating system than runs R.^[Notice that Jake's reproduction archives and/or
instructions for using them are hosted on the
[Dataverse](http://thedata.org/book/learn-about-project), which is another
system designed to enhance academic collaboration across time and space.]  The
idea that one ought to be able to install and run and use an academic paper
just as one installs and uses statistical software packages is very attractive,
and we anticipate that it will become ever easier to turn papers into R
packages as creative and energetic folks turn their attention to the question
of reproducible research.^[See for example <http://r-pkgs.had.co.nz/>] Finally,
(5) Maarten has been using Dropbox for most of his collaborative projects. See
comments above. As a tool for making files publicly available Dropbox is less
useful. You can of course make replication files available through creating an
public folder, but you will still need to refer to this folder on a particular
project website (an example is [Maarten's project on conflict and football](
https://www.dropbox.com/sh/wqosq99019mf04d/AAASTgJLkn6lmZJI_1veUK7ca?dl=0)).

@@

We all always collaborate. Many of us collaborate with groups of people at one
moment in time as we race against a deadline. All of us collaborate with
ourselves over time.^[What is a reasonable time-span for which to plan
for self-collaboration on a single idea?  Ask your advisers how long it took
them to move from idea to dissertation to publication.]  The time-frames over
which collaboration are required --- whether among a group of people working
together or within a single scholar's productive life or probably both --- are
much longer than any given version of any given software will easily exist.
Plain text is the exception. Thus, even as we extol version control systems,
one must have a way to ensure future access to them in a form that will still
be around when sentient cockroaches finally join political science departments
(by then dominated by cetaceans after humans are mostly uploads).^[The arrival
of the six-legged social scientists revives Emacs and finally makes Ctrl-a
Ctrl-x Esc-x Ctrl-c a [reasonable key
combination](http://kieran.healy.usesthis.com/). So far, in terms of version
control systems, Jake has used RCS, CVS, bazaar, subversion, and now mostly
uses git.]

@@

Assume we want others to believe us when we say something. More narrowly,
assume we want other people to believe us when we say something about data:
'data' here can be words, numbers, musical notes, images, ideas, etc. The point
is that we are making some claims about patterns in some collection of stuff.
For example, when Jake was invited into the homes and offices of ordinary
people in Chile in 1991, the "stuff" was recordings of conversations that he
had about life during the first year of democracy. Now, it might be easy to
convince others that 'this collection of stuff' is different from 'that
collection of stuff' if those people were looking over our shoulders the whole
time that we made decisions about collecting the stuff and broke it up into
understandable parts and reorganized and summarized it.  Unfortunately, we
can't assume that people are willing to shadow a researcher throughout her
career. Rather, we do our work alone or in small groups and want to convince
other distant and future people to take our analyses and findings seriously.

>>>>

---
title: How to improve your relationship with your future self, V 2.0
author:
- name: Anon 1
  affiliation: Anon 1 Affiliation
- name: Anon 2
  affiliation: Anon 2 Affiliation and Acknowledgements
bibliography: references.bib
csl: american-political-science-review.csl
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    pdf_document:
      keep_tex: true
      number_sections: true
      fig_width: 7
      fig_height: 7
      fig_caption: true
      template: bowersarticletemplate.latex
      pandoc_args: [
      "-M", "secPrefix=\\S",
      "--filter", "/usr/local/bin/pandoc-crossref"
      ]
      md_extensions: +autolink_bare_uris+ascii_identifiers+tex_math_single_backslash
...

@@

If we cannot count on our memories about why we made such and such a design or analysis decision, then what should we do? How can we minimize regret with our past decisions? How can we improve our relationship with our future self?  This essay is a heavily revised and updated version of anonwhitepaper and provides some suggestions for practices that will make reproducible data analysis easy and quick. Specifically, this piece aims to amplify some of what we already ought to know and do, and highlight some current practices, platforms, and possibilities.^[@king1995replication and @nagler1995coding were two of the first pieces introducing these kinds of ideas to political scientists. Now, the efforts to encourage transparency and ease of learning from the data and analyses of others has become institutionalized with the DA-RT inititiave <http://www.dartstatement.org/> @lupia2014openness. These ideas are spreading beyond political science as well @freese2007replication and @asendorpf2013recommendations, <http://osf.io> and <http://www.bitss.org/>.] We aim to provide practical advice about how to *do work* such that one complies with such recommendations as a matter of course and, in so doing, can focus personal regret on bad past decisions that do not have to do with data analysis and the production of scholarly papers.

@@

Literate programming need not go against the principle of modular data analysis
[@nagler1995coding]. Anon1 routinely uses several different files that fulfill
different functions, some of them create \LaTeX code that he can `\input` into
his `main.tex` file, others setup the data, run simulations, or allow him to
record his journeys down blind alleys. Of course, when we have flying cars
running on autopilot, perhaps something other than combining R and Markdown or
LaTeX will make our lives even easier. Then we'll change.

@@

Best practice is to create a file system where you separate folders by function and separate input from output files. For example, below we show a folder structure for a paper where we look at inequality. The paper is written in .tex, data analysis in Stata (that Anon2 prefers, so we see .do and .dta file extensions) in R (the program of choice for Anon1, see the .R extensions). Anon2 likes to make sure that the directory structure in his projects is the same across projects (so, for example, the numbers on the directory names allow for easy default sorting). The file naming with full dates also allows for easy sorting:

@@

Anon1 tends to be less organized than Anon2 and relies on README.md files to tell his future self what is going on in a given directory, a [version control system](http://github.com) to keep track of versions of files, and a [Makefile](http://kbroman.org/minimal_make/) to keep track of relationships between files.

@@

Many rely on Dropbox, 'track changes' in Word, or Google Docs for
collaboration. Those tools tend to require you to communicate with other folks
in your group before you edit existing files. On Dropbox, only one of you can
edit and save a given file at a time. Tracking changes with Word tends to
function best if two people are not editing the same document (merging the
changes between two documents is not simple or builtin). Google Docs allows for
two people to edit the same document at the same time: but you must be online
with a fairly fast internet connection.^[Note to future selves: change this
section every five years or so. Stop saying 'internet' in 2026.] Communication
and trading off editing a file prevents your work (or your colleagues' work)
from getting lost when you both try to save the same file on top of each other. One may also use Dropbox as a kind of server for version
control (though the downside is if your collaborations or project grows large
(in terms of MBs) you need to switch to Dropbox's paid service). Anon2 has been using
this for most of his collaborations. You can copy files from the Dropbox
directory into a local working directory so as to avoid clobbering and then
work on merging changes by hand before copying back to the Dropbox directory
and replacing existing files. You will however need to agree on an iron file
and directory naming rule with your collaborators to ensure that the files and directories are always named the same across machines and versions.

Another approach to organizing work on many files, including files for code and
for writing, is to use a formal [version
control](https://en.wikipedia.org/wiki/Version_control) strategy. As of 2016,
the standard for managing large collaborations with many files is the
[Git](https://en.wikipedia.org/wiki/Git_(software)) system. User friendly free
front-ends for Git currently make the process of learning and using Git very
convenient and integrated into the kind of workflow that your future self will
be proud of: [Open Science Framework](http://osf.io)
[bitbucket](http://bitbucket.com), [github](http://github.com). Anon1 uses
[github](http://github.com) for all of his collaborations with others as well
as with his future self: his collaborators appreciate some of the nice extra
features of github, such as the ability to keep a [shared task
list](https://guides.github.com/features/issues/). Learning about version
control systems takes a bit of time. We suggest, however, it is well worth the
time investment as it will save (lots!) of time later on.

@@

(1) We wrote this paper in the [R markdown](http://rmarkdown.rstudio.com/)
literate programming format using two different text editors (Anon1 used vim)
and (Anon2 used [atom](https://atom.io/)). We organized our files and
collaboration on Github. The source code for this document can be accessed,
downloaded and modified from <https://github.com/anon1/workflow>. We also
used Github Issues to send notes to each other and maintain a task list. The
nice thing about github is that is enables you to make "releases" of a project
which enable you to smooth the reproduction of all of the products of your
research (simulations, data analyses, tables, figures, etc..).  Similar
features are offered by the [Open Science Framework (OSF)](https://osf.io/).
Anon2 recently participated in a workshop on research transparency organized
by the [Berkeley Institute for Transparency in the Social
Sciences](www.bitss.org) in which all [course material](https://osf.io/qh2nr/)
was accessible through OSF.  OSF integrates with Github nicely and a key
benefit is that both systems have been created to remain for a long time
(unlike perhaps your university drive).  (2) For one paper Anon1 simply mixed R
and \LaTeX code in one document (called the "Sweave" format of literate
programming) and then added that document and data into into a compressed
archive [anon1refA]; (3) For another more computing intensive paper
Anon1 assembled a set of files that enabled reproduction of results using the
`make` system [anon1refB]; (4) Anon1 also tried the "compendium"
approach [@gentleman2005reproducible,@gentleman2007statistical] which embeds an
academic paper within the R package system [anon1refC]; The benefit
of the compendium approach is that one is not required to have access to a
command line for `make`: the compendium is downloadable from within R using
`install.packages()` and is viewable using the `vignette()` function in any
operating system than runs R.^[Notice that Anon1's reproduction archives and/or
instructions for using them are hosted on the
[Dataverse](http://thedata.org/book/learn-about-project), which is another
system designed to enhance academic collaboration across time and space.]  The
idea that one ought to be able to install and run and use an academic paper
just as one installs and uses statistical software packages is very attractive,
and we anticipate that it will become ever easier to turn papers into R
packages as creative and energetic folks turn their attention to the question
of reproducible research.^[See for example <http://r-pkgs.had.co.nz/>] Finally,
(5) Anon2 has been using Dropbox for most of his collaborative projects. See
comments above. As a tool for making files publicly available Dropbox is less
useful. You can of course make replication files available through creating an
public folder, but you will still need to refer to this folder on a particular
project website (an example is [Anon2's project on conflict and football](
https://www.dropbox.com/sh/wqosq99019mf04d/AAASTgJLkn6lmZJI_1veUK7ca?dl=0)).

@@

We all always collaborate. Many of us collaborate with groups of people at one
moment in time as we race against a deadline. All of us collaborate with
ourselves over time.^[What is a reasonable time-span for which to plan
for self-collaboration on a single idea?  Ask your advisers how long it took
them to move from idea to dissertation to publication.]  The time-frames over
which collaboration are required --- whether among a group of people working
together or within a single scholar's productive life or probably both --- are
much longer than any given version of any given software will easily exist.
Plain text is the exception. Thus, even as we extol version control systems,
one must have a way to ensure future access to them in a form that will still
be around when sentient cockroaches finally join political science departments
(by then dominated by cetaceans after humans are mostly uploads).^[The arrival
of the six-legged social scientists revives Emacs and finally makes Ctrl-a
Ctrl-x Esc-x Ctrl-c a [reasonable key
combination](http://kieran.healy.usesthis.com/). So far, in terms of version
control systems, Anon1 has used RCS, CVS, bazaar, subversion, and now mostly
uses git.]

@@

Assume we want others to believe us when we say something. More narrowly,
assume we want other people to believe us when we say something about data:
'data' here can be words, numbers, musical notes, images, ideas, etc. The point
is that we are making some claims about patterns in some collection of stuff.
For example, when Anon1 was invited into the homes and offices of ordinary
people in Chile in 1991, the "stuff" was recordings of conversations that he
had about life during the first year of democracy. Now, it might be easy to
convince others that 'this collection of stuff' is different from 'that
collection of stuff' if those people were looking over our shoulders the whole
time that we made decisions about collecting the stuff and broke it up into
understandable parts and reorganized and summarized it.  Unfortunately, we
can't assume that people are willing to shadow a researcher throughout her
career. Rather, we do our work alone or in small groups and want to convince
other distant and future people to take our analyses and findings seriously.
